Mapping short reads, faithfully

### 2020-02-10 manuscript.tex
Change something

\subsection{Estimating the number of duplicates}

The main source of uncertainty in the measurements of mapping quality
estimation is the existence of sequence duplicates. Duplicates are groups
of sequences that are not exactly identical but share high levels of
similarity, likely because they co-evolved together or resulted from
horizontal DNA transfer events. Duplicate sequences are hard to map
reliably because the similarities between the read and the secondary
candidate references are not negligible. At the same time, the exact
probability that the read originated from one of the duplicates is
difficult to quantify in practice. In this section we introduce a method
to estimate the number of duplicates of a sequence. This estimate is
crucial in the whole process of mapping quality computation because the
number of duplicates will be the main determinant of the quality range.

We assume that the true location of the read is a sequence of the genome
that has $N \geq 0$ duplicates. For every nucleotide, the probability that
a duplicate differs from the target is a constant $\mu$, and the
probability that the read differs from the target is a constant $p$. We
further assume that the genome has size $G$ and that its composition is an
equiprobable random mix of the four nucleotides.

For every nucleotide, the probability that the read differs from a given
duplicate is $\lambda = (1-p)\mu + p(1-\mu/3)$. The probability that the
duplicate matches the last $L$ nucleotides of the read is thus $a_L =
(1-\lambda)^L$, and the probability that at least one of the the $N$
duplicates matches them is $1-(1-a_L)^N$. Finally, the probability that
some duplicate matches the last $L$ nucleotides of the read, but none
matches the last $L+1$ is $\ell(L|N) = (1-a_{L+1})^N-(1-a_L)^N.$

@@

In practice, we can compute $L$ with the backward search by extending the
query until there are fewer than two hits in the genome (when there is
only one hit, it is most likely the target). This is not exactly a maximum
likelihood estimate because we should specifically account for the target
as one of the hits. Also, $\mu$ is usually unknown, so we take a
conservative approach and set $\mu$ to a value where the recall is low for
reads in the range 50--150~bp (for MEM seeds we set $\mu = 0.06$).

@@

\subsection{Computing MAPQ}

Based on the estimate of the number of duplicates, we define three
different groups of reads. The mapping quality for each read group
is computed differently because the dominant mapping uncertainty
depends on the number of duplicates.

\subsubsection{Super quality} Super reads are defined as reads that
satisfy the two following conditions: 20-mers extracted every 10
nucleotides all have a single hit in the genome (see the Results section
and Figure~\ref{fig_supertest}), and the read is mapped without mismatch.
Super reads are mapped to an incorrect location iif:
\begin{enumerate}
  \item the read contains $m$ errors
  \item the target has exactly one duplicate
  \item the duplicate differs from the target exactly at the $m$ locations
    of the errors
  \item the $m$ incorrect nucleotides match the sequence of the duplicate
  \item the $m$ errors are in the odd-numbered 10-mers of the read
\end{enumerate}
The justification for these conditions are explained in more detail in
Figure~\ref{fig_supread}.

The last condition imposes $m = 1 + \lfloor L/20 \rfloor$ for a read of size
$L$. The probability of condition $1$ is denoted $\alpha$. The probability
of condition $2$ is ${L \choose m}p^m(1-p)^{L-m}$, where $p$ is the error
rate of the read sequencer. The probability of condition $3$ is
$\mu^m(1-\mu)^{L-m}$, where $\mu$ is the probability that the duplicate
differs from the target. The probability of condition $4$ is $1/3^m$ and the
probability of condition $5$ is $10^m/{L\choose m}$.

We need to multiply the probability of each condition together and divide
them by the probability that a read is a \emph{super read}, which is fraction
$\beta$ of the probability that a read has no error. The mapping quality of
super reads is finally computed as
\begin{align}
\label{eq_super}
\frac{\alpha}{\beta} \left( \frac{3.33 \cdot p\mu}{1-p}
\right)^m (1-u)^{L-m}.
\end{align}

@@

\subsubsection{Normal quality} Reads with initial estimate $N \leq 20$ are
mapped by checking all the candidates. The prior probability that the true
location is not discovered can be calculated with
\texttt{sesame}~\cite{Filion619155}. We can gain precision by computing the
posterior probability given the number of mismatches (intuitively, a large
number of mismatches is evidence that the read is mapped to an incorrect
location). If the read is mapped to the correct location, the number of
errors follows a binomial distribution where the parameter $p$ is the
error rate of the sequencer. If it is mapped to a paralog, the parameter
is $\lambda$ as defined above. A complication is that the distribution of
errors is not binomial when $N > 1$ because the best hit is not chosen
randomly among the $N$ paralogs. We thus compute the posterior probablity
using Bayes formula, but put an upper threshold equal to 1 on the
log-ratio of the evidence. This means that many mismatches is strong
support for an incorrect location, but few mismatches is only weak support
for the correct location.

@@

It was verified empirically that expression (\ref{eq_bayes}) is an
overestimate when $x = 0$. In this case, the estimate is replaced by a
direct approach where we compute the probability that a paralog matches
the read perfectly. More specifically, we compute the probability that the
read has exactly one error and that a paralog matches the incorrect
sequence. In this case, the probability that the location is incorrect is
computed as
\begin{align}
\label{eq_x0}
P(N \geq 1) \cdot \frac{Lp(1-p)^{L-1}
\left(1-\left((1-\mu)^{L-1}\mu/3\right)^N\right)}{(1-p)^L}.
\end{align}

@@

\subsubsection{Low quality} Reads with initial estimate $N > 20$ are
mapped by checking only the longest MEM seed. A seed of size $S$ exists if
there are $S$ nucleotides without error or if there are $S$ nucleotides
with possibly many errors matching at least one of the $N$ paralogs. In the
first case, the target is discovered, in the second it is not. We also have
to account whether the seed is flanked by errors or by the end of the read.
The probability of the first case is
\begin{align*}
P_{\text{hit}} = \delta_L \cdot (1-p)^S \cdot \delta_R,
\end{align*}
where $\delta_{L/R} = 1$ if the seed is flanked by the end of the read and
$p$ otherwise.

@@

We denote $m$ the longest query size with two more more hits in the genome
(see Figure~\ref{fig_back}) and we write $P(m)$ in the three cases above.
The genome has size $G$ and every nucleotide matches the query with
probability $1/4$ if it is unrelated to the target, and with probability
$\lambda$ if it is a paralog. For convenience, we also define the
probability of a random hit of size less than $m$ as $\xi(m) = 1-1/4^m$.

@@

Finally, to compute the probability of case $i.$ given the observed value
of $m$, we use Bayes formula with prior odds in proportion $1:1:8$ because
we estimate that the probability that a sequence has exactly one paralog
is approximately $1/10$ (remember that those calculations are performed
when there is evidence that $N \leq 20$).

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.6]{MEM_19.pdf}
\end{center}
\caption{Projected performance of MEM seeds. The probability that the
target is not in the candidate set after seeding  is plotted against the
size of the read (notice the log scale on the $y$ axis). Computations are
carried out with \texttt{sesame} for MEM seeds of minimum size 19,
assuming an error rate of 1\% and a divergence rate of 6\% between
paralogs (the probability that a given nucleotide of the target is
different in a given paralog is 0.06). Each line represents the
probabilities for a different number $N$ of paralogs. Note that the
asymptotic decay is approximately constant up to $N = 20$ and slows down
for larger $N$.}
\label{fig_MEM}
\end{figure}

The results reveal two essential insights: The first is that the
asymptotic decay of the failure rate is approximately constant when the
read maps to a location with 20 or fewer duplicates. Beyond this, the
asymptotic decay slows down and the recall decreases markedly. The second
insight is that up to 100 nucleotides, the failure rate is always above 1
error per 100,000 reads. In other words, there is no hope to map a read
with very high confidence using MEM seeds if the target has a paralog.

These insights suggest that an efficient strategy is to use a fast mapping
mode when the target has many duplicates because the read cannot be mapped
with high confidence anyway. In contrast, when the number of duplicates is
low, it is important to know whether the target is a unique sequence
because this is the only way to map a read with high confidence. So after
MEM seeding, our strategy is to quickly estimate the number of duplicates,
find a decent candidate location as fast as possible and finally estimate
how reliable this location is. In essence, we search the reads that
can be mapped with high confidence instead of searching a high-confidence
location for every read. A high-level summary of this strategy is shown in
Figure~\ref{fig_tree}.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.8]{decision_tree.pdf}
\end{center}
\caption{Proposed strategy for a faithful mapper. Each read is routed
through a different mapping pathway. First, all the MEM seeds are
extracted (top). Then the number $N$ of paralogs of the target is
estimated. If $N$ is 20 or less (left branch), all the candidates are
tested with the Needleman-Wunsch algorithm. Once the best candidate is
found, the probability that it is unique is estimated. If it is high and
the alignment has no mismatch, mapping quality is computed using
(\ref{eq_super}). If the best candidate is not unique or if the alignment
has a mismatch, mapping quality is computed using (\ref{eq_bayes}) or
(\ref{eq_x0}). If $N$ is higher than 20 (right branch), only the largest
MEM seed is used and the mapping quality is computed using (\ref{eq_low}).
}
\label{fig_tree}
\end{figure}

@@

The key in the flow chart of Figure~\ref{fig_tree} is to dispatch the
reads based the number $N$ of paralogs of the target, which is an
indirect measure of the mappability of the read~\cite{pmid22276185}.
Estimating $N$ must be fast so that this step does not become a
bottleneck. To achieve this, we propose to use the seeding process itself.
MEM seeds are best computed using the backward
search~\cite{ferragina2000opportunistic}, an algorithm that returns the
number of hits in the genome as the query is extended backward. If the
target has no or few paralogs (Figure~\ref{fig_back}a), the number of hits
will quickly go below 2 (when there is only one hit, it is most likely the
target itself). In contrast, if the target has many paralogs in the genome
(Figure~\ref{fig_back}b), the number of hits will remain at 2 or above for
many additional iterations.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.84]{pushing_backward.pdf}
\end{center}
\caption{Estimating $N$ with the backward search. \textbf{(a)} The target
has no or few paralogs. It takes only a few iterations until the backward
search (BWS) finds fewer than 2 hits in the genome. \textbf{(b)} the
target has many paralogs. In this case it takes more iteration before the
query has fewer than 2 hits. The number of iterations above the threshold
is used as a statistic to estimate $N$.}
\label{fig_back}
\end{figure}

@@

We designed another estimation procedure based on the backward search,
where this time we use the best candidate location and not the read as a
query. The principle, sketched in Figure~\ref{fig_supertest}a, is to
perform the backward search on 20-mers and to test whether the genome
contains a single hit. The rationale is that the
total number of 20-mers is approximately $10^{12}$, so the chances of
spurious random hits are very low, even in the largest known genomes.

@@

The extremely high mapping quality of super reads comes from an unexpected
property of the test depicted in Figure~\ref{fig_supertest}. To see why
they can be mapped with such confidence, consider a read aligning
perfectly to an incorrect sequence of the genome. Observe that for
this to happen, the read must contain one or more errors that are perfectly
compensated in some duplicate of the target. This can only happen if $i.$
the target has a duplicate, $ii.$ the read contains an error, $iii.$ this
error matches the duplicate and $iv.$ the target and the duplicate are
otherwise identical.

@@

But those conditions are not sufficient to make a read \emph{super}. As
shown in Figure~\ref{fig_supread}a, the mapping location would not be
considered unique because there would exist at least one 20-mer where the
target and its dupliate are identical. For a read of size 50, the
duplicate can be overlooked only if there are at least two errors in the
read that match the paralog.

But as shown in Figure~\ref{fig_supread}b, this is still not sufficient
for a wrongly mapped read to qualify as \emph{super}. For this, every
second 10-mer must contain an error that is exactly compensatd in the
duplicate, as indicated in Figure~\ref{fig_supread}c. In this
configuration, the errors mask the presence of a duplicate on all the
tested 20-mers and the incorrect target location is considered unique.
There are other scenarios where a super read is mapped to an incorrect
location, but they involve even more compensated errors so they can be
neglected.

@@

The present work focuses on general mapping tasks in eukaryotic genomes,
but faithful read mappers may find much more interesting applications,
such as when the reference contains several variants of the same
sequences. This is the case for the genome of hybrid species or of
heterozygote individuals where each sequence has $N \geq 1$ duplicates in
the reference. In this case, the challenge is to estimate the confidence
that the read maps to one genome or the other. MMP could be used for this
kind of problem, but a more specialized method that capitalizes on the key
information $N \geq 1$ would be more appropriate. For instance, the time
spent looking for super reads is wasted in this context because they
cannot occur (not a single sequence is unique). A more specialized method
dedicated to calibrating low mapping quality is expected to give better
results.

Another case of interest is when the DNA sample is contaminated and needs
to be mapped to several species. This occurs when working with primates
because experimenters can contaminate the biological material with their
own DNA, or simply when the source of the DNA is unknown. In such cases,
only few sequences are expected to have no duplicate in the reference. In
particular, if the genomes of interest are relatively close, the copy
number will usually be equal to the number of genomes. In such cases, a
general faithful mapper such as MMP is expected to give decent results,
even though it was not tested in this context. More generally, we expect
faithful mapping algorithms to find other applications in many areas of
biology.

>>>>

\subsection{Estimating the number of paralogs}

The main source of uncertainty in the mapping is the existence of
near-repetitive sequences, here referred to as paralogs. Paralogs are
groups of sequences that share high levels of similarity, likely because
they resulted from duplication or horizontal DNA transfer events. Paralogs
are hard to map reliably because the read can be similar to some secondary
candidates. At the same time, the exact probability that the read
originated from one of the paralogs is difficult to quantify in practice.
In this section we introduce a method to estimate the number of paralogs
of a sequence. This estimate is crucial in the whole process of computing
the mapping quality because the number of paralogs will be the main
determinant of the quality.

We assume that the true location of the read is a sequence of the genome
that has $N \geq 0$ paralogs. For every nucleotide, the probability that a
paralog differs from the target is a constant $\mu$ (this event is a
mutation), and the probability that the read differs from the target is a
constant $p$ (this event is a sequencing error). We further assume that
the genome has size $G$ and that its composition is an equiprobable random
mix of the four nucleotides.

For every nucleotide, the probability that the read differs from a given
paralog is $\lambda = (1-p)\mu + p(1-\mu/3)$. The probability that the
paralog matches the last $L$ nucleotides of the read is thus $a_L =
(1-\lambda)^L$, and the probability that at least one of the the $N$
paralogs matches them is $1-(1-a_L)^N$. Finally, the probability that
some paralog matches the last $L$ nucleotides of the read, but none
matches the last $L+1$ is $\ell(L|N) = (1-a_{L+1})^N-(1-a_L)^N.$

@@

In practice, we can compute $L$ with the backward search by extending the
query until there are fewer than two hits in the genome (when there is
only one hit, it is most likely the target). This is not exactly a maximum
likelihood estimate because we should specifically account for the target
as one of the hits. Also, $\mu$ is usually unknown, so we take a
conservative approach and set $\mu$ to a value where the recall is lowest
for reads in the range 50--150~bp (for MEM seeds we set $\mu = 0.06$).

@@

\subsection{Mapping quality}

Based on the estimate of the number of paralogs, we define three
different groups of reads. The mapping quality for each read group
is computed differently because the dominant mapping uncertainty
depends on the number of paralogs.

\subsubsection{Super quality} Super reads are defined as reads that
satisfy the two following conditions: 20-mers extracted every 10
nucleotides all have a single hit in the genome (see the Results section
and Figure~\ref{fig_supertest}), and the read is mapped without mismatch.
Super reads are mapped to an incorrect location if:
\begin{enumerate}
  \item the read contains $m$ errors,
  \item the target has exactly one paralog,
  \item the paralog differs from the target exactly at the $m$ locations
    of the errors,
  \item the $m$ incorrect nucleotides match the sequence of the paralog,
  \item the $m$ errors are in the odd-numbered 10-mers of the read.
\end{enumerate}
The justification for these conditions are explained in more detail in
Figure~\ref{fig_supread}.

The last condition imposes $m = 1 + \lfloor L/20 \rfloor$ for a read of
size $L$. The probability of condition $1$ is denoted $\alpha$. The
probability of condition $2$ is ${L \choose m}p^m(1-p)^{L-m}$, where $p$
is the error rate of the sequencer. The probability of condition $3$
is $\mu^m(1-\mu)^{L-m}$, where $\mu$ is the probability that the paralog
differs from the target. The probability of condition $4$ is $1/3^m$ and
the probability of condition $5$ is $10^m/{L\choose m}$.

We need to multiply the probability of each condition together and divide
them by the probability that a read is a \emph{super read}, which is a
fraction $\beta$ of the probability that a read has no error. The mapping
quality of super reads is finally computed as
\begin{align}
\label{eq_super}
\frac{\alpha}{\beta} \left( \frac{3.33 \cdot p\mu}{1-p}
\right)^m (1-u)^{L-m}.
\end{align}

@@

\subsubsection{Normal quality} Reads with initial estimate $N \leq 20$ are
aligned to all the locations. The prior probability that the true location
was not discovered at the seeding step can be calculated with
\texttt{sesame}~\cite{Filion619155}. We can gain precision by computing
the posterior probability given the number of mismatches (intuitively, a
large number of mismatches is evidence that the read is mapped to an
incorrect location). If the read is mapped to the correct location, the
number of errors follows a binomial distribution where the parameter $p$
is the error rate of the sequencer. If it is mapped to a paralog, the
parameter is $\lambda$ as defined above. A complication is that the
distribution of errors is not binomial when $N > 1$ because the best hit
is not chosen randomly among the $N$ paralogs. We thus compute the
posterior probablity using Bayes' formula, but put an upper threshold equal
to 1 on the log-ratio of the evidence. This means that many mismatches
give strong support for an incorrect location, but few mismatches give
only weak support for the correct location.

@@

It was verified empirically that expression (\ref{eq_bayes}) is an
overestimate when $x = 0$ (because here the upper threshold of 1 is too
pessimistic). In this case, the estimate is replaced by a direct approach
where we compute the probability that a paralog matches the read
perfectly. More specifically, we compute the probability that the read has
exactly one error and that a paralog matches the incorrect sequence. In
this case, the probability that the location is incorrect is computed as
\begin{align}
\label{eq_x0}
P(N \geq 1) \cdot \frac{Lp(1-p)^{L-1}
\left(1-\left((1-\mu)^{L-1}\mu/3\right)^N\right)}{(1-p)^L}.
\end{align}

@@

\subsubsection{Low quality} Reads with initial estimate $N > 20$ are
aligned only to the longest MEM seed. A seed of size $S$ exists if
there are $S$ nucleotides without error or if there are $S$ nucleotides
with possibly many errors matching at least one of the $N$ paralogs. In the
first case, the target is discovered, in the second it is not. We also have
to know whether the seed is flanked by errors or by the end of the read.
The probability of the first case is
\begin{align*}
P_{\text{hit}} = \delta_L \cdot (1-p)^S \cdot \delta_R,
\end{align*}
where $\delta_{L/R} = 1$ if the seed is flanked by the end of the read and
$\delta_{L/R} = p$ otherwise.

@@

We denote $m$ the longest query size with two or more hits in the genome
(see Figure~\ref{fig_back}) and we work the probability $P(m)$ in the
three cases above. The genome has size $G$ and every nucleotide matches
the query with probability $1/4$ if it is unrelated to the target, and
with probability $\lambda$ if it is a paralog. For convenience, we also
denote the probability of a random hit of size less than $m$ as $\xi(m) =
1-1/4^m$.

@@

Finally, to compute the probability of case $i.$ given the observed value
of $m$, we use Bayes' formula with prior odds in proportion $1:1:8$ because
we estimate that the probability that a sequence has exactly one paralog
is approximately $1/10$ (remember that those calculations are performed
when there is evidence that $N \leq 20$).

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.6]{MEM_19.pdf}
\end{center}
\caption{Projected performance of MEM seeds. The probability of filtering
out the target is plotted against the size of the read (notice the log
scale on the $y$ axis). Computations are carried out with \texttt{sesame}
for MEM seeds of minimum size 19, assuming an error rate of 1\% and a
divergence rate of 6\% between paralogs (\textit{i.e.} the probability
that a given nucleotide of the target is different in a given paralog is
0.06). Each line represents the probabilities for a different number $N$
of paralogs. Note that the asymptotic decay is approximately constant up
to $N = 20$ and slows down for larger $N$.}
\label{fig_MEM}
\end{figure}

The results reveal two essential insights: The first is that the
asymptotic decay of the failure rate is approximately constant when the
read maps to a location with $N \leq 20$ paralogs. Beyond this, the
asymptotic decay slows down and the recall decreases markedly. The second
insight is that up to 100 nucleotides, the failure rate is always above 1
mapping error per 100,000 reads. In other words, there is no hope to map a
read with very high confidence using MEM seeds if the target has a
paralog.

These insights suggest that an efficient strategy is to use a fast mapping
mode when the target has many paralogs because the read cannot be mapped
with high confidence anyway. In contrast, when the number of paralogs is
low, it is important to know whether the target is a unique sequence
because this is the only way to map a read with high confidence. So after
MEM seeding, our strategy is to quickly estimate the number of paralogs,
find a decent candidate location as fast as possible and finally estimate
how reliable this location is. In essence, we search the reads that
can be mapped with high confidence instead of searching a high-confidence
location for every read. A high-level summary of this strategy is shown in
Figure~\ref{fig_tree}.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.8]{decision_tree.pdf}
\end{center}
\caption{Proposed strategy for a faithful mapper. Each read is routed
through one of three mapping pathways. First, all the MEM seeds are
extracted (top). Then the number $N$ of paralogs of the target is
estimated. If $N$ is 20 or less (left branch), all the candidates are
tested with the Needleman-Wunsch algorithm. Once the best candidate is
found, the probability that it is unique is estimated. If it is high and
the alignment has no mismatch, mapping quality is computed using
expression (\ref{eq_super}). If the best candidate is not unique or if the
alignment has a mismatch, mapping quality is computed using expression
(\ref{eq_bayes}) or (\ref{eq_x0}). If $N$ is higher than 20 (right
branch), only the largest MEM seed is used and the mapping quality is
computed using expression (\ref{eq_low}).}
\label{fig_tree}
\end{figure}

@@

The key in the flow chart of Figure~\ref{fig_tree} is to dispatch the
reads based on the number $N$ of paralogs of the target, which is an
indirect measure of the mappability of the read~\cite{pmid22276185}.
Estimating $N$ must be fast so that this step does not become a
bottleneck. To achieve this, we propose to use the seeding process itself.
MEM seeds are best computed using the backward
search~\cite{ferragina2000opportunistic}, an algorithm that returns the
number of hits in the genome as the query is extended backward. If the
target has no or few paralogs (Figure~\ref{fig_back}a), the number of hits
will quickly go below 2 (when there is only one hit, it is most likely the
target itself). In contrast, if the target has many paralogs
(Figure~\ref{fig_back}b), the number of hits will remain at 2 or above for
many additional iterations.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.84]{pushing_backward.pdf}
\end{center}
\caption{Estimating $N$ with the backward search. \textbf{(a)} The target
has no or few paralogs. It takes only a few iterations until the backward
search (BWS) finds fewer than 2 hits in the genome. \textbf{(b)} The
target has many paralogs. In this case it takes more iterations before the
query has fewer than 2 hits. The number of iterations above the threshold
is used as a statistic to estimate $N$.}
\label{fig_back}
\end{figure}

@@

We designed another estimation procedure based on the backward search,
where this time we use the best candidate location as a query instead of
the read. The principle, sketched in Figure~\ref{fig_supertest}a, is to
perform the backward search on 20-mers and to test whether the genome
contains a single hit. The rationale is that the total number of 20-mers
is approximately $10^{12}$, so the chances of spurious random hits are
very low, even in the largest known genomes.

@@

The extremely high mapping quality of super reads comes from an unexpected
property of the test depicted in Figure~\ref{fig_supertest}. To see why
they can be mapped with such confidence, consider a read aligning
perfectly to an incorrect sequence of the genome. Observe that for
this to happen, the read must contain one or more errors that are perfectly
compensated in some paralog of the target. This can only happen if $i.$
the target has a paralog, $ii.$ the read contains an error, $iii.$ this
error matches the paralog and $iv.$ the target and the paralog are
otherwise identical.

@@

But those conditions are not sufficient to make a read \emph{super}. As
shown in Figure~\ref{fig_supread}a, the mapping location would not be
considered unique because there would exist at least one 20-mer where the
target and its dupliate are identical. For a read of size 50, the
paralog can be overlooked only if there are at least two errors in the
read that match the paralog.

But as shown in Figure~\ref{fig_supread}b, this is still not sufficient
for a wrongly mapped read to qualify as \emph{super}. For this, every
second 10-mer must contain an error that is exactly compensatd in the
paralog, as indicated in Figure~\ref{fig_supread}c. In this
configuration, the errors mask the presence of a paralog on all the
tested 20-mers and the incorrect target location is considered unique.
There are other scenarios where a super read is mapped to an incorrect
location, but they involve even more compensated errors so they can be
neglected.

@@

The present work focuses on general mapping tasks in eukaryotic genomes,
but faithful read mappers may find much more interesting applications,
such as when the reference contains several variants of the same
sequences. This is the case for the genome of hybrid species or of
heterozygote individuals where each sequence has $N \geq 1$ paralogs in
the reference. In this case, the challenge is to estimate the confidence
that the read maps to one genome or the other. MMP could be used for this
kind of problem, but a more specialized method that capitalizes on the key
information $N \geq 1$ would be more appropriate. For instance, the time
spent looking for super reads is wasted in this context because they
cannot occur (not a single sequence is unique). A more specialized method
dedicated to calibrating low mapping quality is expected to give better
results.

Another case of interest is when the DNA sample is contaminated and needs
to be mapped to several species. This occurs when working with primates
because experimenters can contaminate the biological material with their
own DNA, or simply when the source of the DNA is unknown. In such cases,
only few sequences are expected to have no paralog in the reference. In
particular, if the genomes of interest are relatively close, the copy
number will usually be equal to the number of genomes. In such cases, a
general faithful mapper such as MMP is expected to give decent results,
even though it was not tested in this context. More generally, we expect
faithful mapping algorithms to find other applications in many areas of
biology.
